# Case Study

To illustrate some of these ideas we can use a case study of influenza in a boarding school. These data are from a paper in the BMJ in 1978 [@anonymous:1978] and provided in the [`outbreaks`](http://www.repidemicsconsortium.org/outbreaks/index.html) R package. We use a simple $SIRR_1$ model with two removal classes, $R$ and $R_1$. We will assume that the bed-rest counts in the data correspond to the number of individuals in the $R$ class, and will ignore the other time-series for the time being. This has been used various times in the literature, including @murray:2003, @devriesetal:2006 and in some of the [`pomp` tutorials](https://kingaa.github.io/sbied/mif/mif.html#an-iterated-filtering-algorithm-if2). The stochastic model we will use has event probabilities:
\begin{align*}
    P\left[S_{t + \delta t} = S_t - 1, I_{t + \delta t} = I_t + 1\right] &\approx \beta S I / N\\
    P\left[I_{t + \delta t} = I_t - 1, R_{t + \delta t} = R_t + 1\right] &\approx \gamma I\\
    P\left[R_{t + \delta t} = R_t - 1, R_{1, t + \delta t} = R_{1,t} + 1\right] &\approx \gamma_1 R
\end{align*}
The initial population size is 763 pupils, and we assume an initial introduction of infection of a single individual at day 0.

```{r, case-BS, message = F, warning = F}
## load data
library(outbreaks)
flu <- influenza_england_1978_school
flu$day <- 1:nrow(flu)
```

```{r, plot-BS, message = F, warning = F, fig.height = 3.5, fig.width = 3.5}
## plot data
plot(flu$day, flu$in_bed, xlab = "Day", ylab = "Number in bed-rest", type = "l")
```
