# Case Study

To illustrate some of these ideas we can use a case study of influenza in a boarding school. These data are from a paper in the BMJ in 1978 [@anonymous:1978] and provided in the [`outbreaks`](http://www.repidemicsconsortium.org/outbreaks/index.html) R package. Following @murray:2003, we will assume that bed-rest event correspond to the onset-of-infectiousness, with the event probabilities are:
\begin{align*}
    P\left[\mbox{infection event in}~[t, t + \delta t)\right] &= \beta S I / N + o(\delta t)\\
    P\left[\mbox{removal event in}~[t, t + \delta t)\right] &= \gamma I + o(\delta t)
\end{align*}
Here we are going to match to the final epidemic size (512 children), and the sum-of-squared differences between the observed infection ("bed-rest") counts and the simulated infection counts (given by the $I$ class in our model). The initial population size is 763 pupils, and we assume an initial introduction of infection of a single boy at day 0.

```{r, case-BS, message = F, warning = F}
## load data
library(outbreaks)
flu <- influenza_england_1978_school
flu$day <- 1:nrow(flu)
```

```{r, plot-BS, message = F, warning = F, fig.height = 3.5, fig.width = 3.5}
## plot data
plot(flu$day, flu$in_bed, xlab = "Day", ylab = "Number of infections", type = "l")
```
