# Set up simple simulation model

This chapter is to illustrate how to use `mparseRcpp` to simulate from a simple epidemic model. Here we will use the model that we wish to fit to the boarding school influenza data.

```{r, sim-simbiid, warning = F, message = F}
## load library
library(SimBIID)
```

To set up the model above we need a character vector of transition rates, as per `mparse()`:

```{r, sim-trans}
## test compilation model
transitions <- c(
    "S -> beta * S * I / (S + I + R + R1) -> I", 
    "I -> gamma * I -> R",
    "R -> gamma1 * R -> R1"
)
```

We also need a vector of compartment names and parameter names (note that we do not set values for the parameters here---instead we pass these and the initial states into later functions):

```{r, sim-comp}
compartments <- c("S", "I", "R", "R1")
pars <- c("beta", "gamma", "gamma1")
```

We then build a model using the `mparseRcpp()`:

```{r, sim-model}
model <- mparseRcpp(
    transitions = transitions, 
    compartments = compartments,
    pars = pars
)
```

The `SimBIID` package provides a `run()` method to run individual or multiple simulations from the model. By default the `mparseRcpp` creates a function that takes four arguments:

* `pars`: a vector of parameter values;
* `tstart`: the time to begin the simulation;
* `tstop`: the time to end the simulation;
* `u`: states of the system at `tstart`.

We can then pass this object to the `run()` method, along with the corresponding arguments e.g.

`r ifelse(opts_knit$get("rmarkdown.pandoc.to") == "latex", "\\newpage", "")`

```{r, sim-seed, include = F}
set.seed(1)
```

```{r, sim-run}
## define parameters
simPars <- c(beta = 2.5, gamma = 1, gamma1 = 1)

## define initial states
iniStates <- c(S = 762, I = 1, R = 0, R1 = 0)

## run model
sims <- run(
    model = model,
    pars = simPars,
    tstart = 0,
    tstop = 15,
    u = iniStates
)
sims
```

Currently this outputs a `tibble` object (essentially a `data.frame`), with columns: `completed / t / u*`, where `completed` is a binary variable containing the value `1` if the epidemic finished before time point `tstop`, and `0` otherwise. Then the value `t` is the time point when the epidemic finished (or `tstop` if the epidemic was still ongoing when the function exited), and I'm using `u*` generically to correspond to a copy of the states of the system at the final time point (here $S$, $I$, $R$ and $R_1$).

Here we can see that the simulation finished at time point `t = `r round(sims$sums[3], 2)``, with the final state of the system given by ``r paste0("c(S = ", sims$sums[4], ", I = ", sims$sums[5], ", R = ", sims$sums[6], ", R1 = ", sims$sums[7], ")")``.

## Using `tspan`

Alternatively, we may want to return the states of the system at a series of pre-determined points. We can do this by adding a `tspan` argument, which is a vector of time points at which we wish to return the states. We initialise the model by telling `mparseRcpp` that we wish to include a `tspan` argument e.g.

```{r, sim-mod1}
## generate model
model <- mparseRcpp(
    transitions = transitions, 
    compartments = compartments,
    pars = pars,
    tspan = T
)
```

When we `run()` the model, we can enter a suitable `tspan` argument as a vector of time points at which we wish to return the states e.g.

`r ifelse(opts_knit$get("rmarkdown.pandoc.to") == "latex", "\\newpage", "")`

```{r, sim-seed1, include = F}
set.seed(10)
```

```{r, sim-run1}
sims <- run(
    model = model,
    pars = simPars,
    tstart = 0,
    tstop = 15,
    u = iniStates,
    tspan = 1:14
)
sims
```

Here we can see that the epidemic `r ifelse(sims$sums[2] == 1, "died out", "was still going")` at time point `t = `r round(sims$sums[3], 1)``.

`r ifelse(opts_knit$get("rmarkdown.pandoc.to") == "latex", "\\newpage", "")`

## Plotting simulations

The `run()` method outputs a `SimBIID_runs` object, which can be plotted using the `plot()` function:

```{r, sim-plot, fig.width = 4, fig.height = 3}
## plot simulation
plot(sims)
```

## Running multiple simulations

We can run multiple simulations by passing a `nrep` argument to `run()`. For example, to run 100 replicates and plot them:

```{r, sim-run2}
sims <- run(
    model = model,
    pars = simPars,
    tstart = 0,
    tstop = 15,
    u = iniStates,
    tspan = 1:14,
    nrep = 100
)
```

`r ifelse(opts_knit$get("rmarkdown.pandoc.to") == "latex", "\\newpage", "")`

```{r, sim-run21, fig.width = 4, fig.height = 3}
sims
plot(sims, quant = c(0.6, 0.7, 0.8, 0.9))
```

Here the ribbons correspond to different prediction bands, and the red line is the posterior median. We can also add individual trajectories by inputing their replicate number, for example below we add the trajectories for replicates 1 and 2 to the plot.

`r ifelse(opts_knit$get("rmarkdown.pandoc.to") == "latex", "\\newpage", "")`

```{r, sim-plot1, fig.width = 4, fig.height = 3}
plot(sims, rep = c(1, 2), quant = c(0.6, 0.7, 0.8, 0.9))
```

There is an argument in `run()` to enable parallel processing if required (if the `parallel` package is installed). See help file for `run()` (e.g. `?run`) for more details.
