# Approximate Bayesian Computation

Here we will run the ABC-SMC routine of @tonietal:2009. Firstly, load the `SimBIID` library:

```{r, abc-simbiid, warning = F, message = F}
## load library
library(SimBIID)
```

```{r, abc-sums, include = F}
## load data
BS <- read.csv("BS.csv", header = T)

## extract summaries
finaltime <- max(BS$day)
finalsize <- 512
```

> **Note: in all the following examples I have used a low number of particles to speed things up. In practice you would want to use many more to ensure better approximations.**

## Simulation model

In order to use the any ABC routine, we require a simulation model. Here we will use the approach described in the previous chapter, and specify a model using the `mparseRcpp()` function. 

```{r, abcsmc-model}
## setup simulation model
transitions <- c(
    "S -> beta * S * I / (S + I + B + C) -> I", 
    "I -> muI * I -> B", 
    "B -> muB * B -> C"
)
compartments <- c("S", "I", "B", "C")
pars <- c("beta", "muI", "muB")
model <- mparseRcpp(
    transitions = transitions, 
    compartments = compartments,
    pars = pars
)
```

## Arguments to `ABCSMC()` function

The ABC-SMC routine is coded in a function called `ABCSMC()`. If you look at the help file for the `ABCSMC()` function (e.g. `?ABCSMC`), you will see the main arguments to the `ABCSMC()` function , which are summarised below:

* `x`: A named vector with entries containing the observed summary statistics to match to. Names must match to `tols` argument.
* `priors`: A `data.frame` containing columns: `parnames`, `dist`, `p1` and `p2`, with number of rows equal to the number of parameters. The column `parname` simply gives names to each parameter for plotting and summarising. Each entry in the `dist` column must contain one of `c("unif", "norm", "gamma")`, and the corresponding `p1` and `p2` entries relate to the hyperparameters (lower and upper bounds in the uniform case; mean and standard deviation in the normal case; and shape and rate in the gamma case).
* `func`: Function that runs the simulator and checks whether the simulation matches the data. The first four arguments must be `pars`, `data`, `tols` and `u`. If the simulations do not match the data then the function must return an `NA`, else it must returns a vector of simulated summary measures. In this latter case the output from the function must be a vector with length equal to `ncol(data)` and with entries in the same order as the columns of `data`.
* `u`:A named vector of initial states.
* `tols`: A vector or matrix of tolerances, with the number of rows defining the number of generations required, and columns defining the summary statistics to match to. If a vector, then the length determines the summary statistics. The columns/entries must match to those in `x`.
* `ptols`: The proportion of simulated outcomes at each generation to use to derive adaptive tolerances.
* `ngen`: The number of generations of ABC-SMC to run.
* `npart`: An integer specifying the number of particles.

### Summary statistics

The first argument to `ABCSMC()` must be either an `ABCSMC` object (we will come to this later), or a named `vector` with entries containing the observed summary statistics to match to. The data here are time-series counts of ***bed-rest *** and ***convalescence*** events. A very simple option would be to match to:

* final epidemic size (i.e. total number of removals across the time course of the epidemic), and
* time of final removal (in this case when the epidemic process ceased).

Although simple, these two measures serve to give us some information on both the **length** and **magnitude** of the epidemic, and should contain at least some useful information about the parameters. In this case the final removal time is `r finaltime` days and the final epidemic size is `r finalsize` individuals. Let's create this data object:

```{r, abc-sums1}
## load data
BS <- read.csv("BS.csv", header = T)

## create vector of summary statistics
sumStat <- c(finalsize = 512, finaltime = 14)
sumStat
```

### Prior distributions {#priors}

The second argument to the `ABCSMC()` function must be a `data.frame` containing information about the prior distributions for the parameters. Here we will let $\beta \sim U(0, 5)$, $\mu_I \sim U(0, 5)$ and $\mu_B \sim U(0, 1)$. We can specify this as follows:

`r ifelse(opts_knit$get("rmarkdown.pandoc.to") == "latex", "\\newpage", "")`

```{r, abc-priors}
## set priors
priors <- data.frame(parnames = c("beta", "muI", "muB"), 
                     dist = rep("unif", 3), 
                     stringsAsFactors = F)
priors$p1 <- c(0, 0, 0)
priors$p2 <- c(5, 5, 1)
priors
```

Here the first column corresponds to the names of the parameters, the second to the type of prior distribution (`"unif"`, `"gamma"` or `"norm"`), and the final two columns to the hyperparameters (upper and lower limits for `"unif"`; shape and rate for `"gamma"`; or mean and standard deviation for `"norm"`).

### Simulation function

The third argument to `ABCSMC()` is a function that runs the simulator and checks whether the simulation matches the data. The first four arguments to this function must be `pars`, `data`, `tols` and `u` (but you can pass more arguments in if required---see below). Within the function:

* `pars`: is a *vector* of parameters (the order must match the order of the `priors` argument to the `ABCSMC()` function);
* `data`: a *vector* of data points to match to (must match the order of the `x` argument to the `ABCSMC()` function);
* `tols`: a *vector* of tolerances (must match the order of `data`);
* `u`: a *vector* of initial states (must match order of the `u` argument to the `ABCSMC()` function).

> **Please note: there is no internal check on you getting these orders correct. It is up to you to be careful when setting up this function!**

If the simulations do not match the data then the function must return an `NA`, else it must returns a vector of simulated summary measures. In this latter case the output from the function must be a vector with length equal to `length(data)` and with entries in the same order as `data`. It is possible to add further arguments to this function, as long as they appear after `pars`, `data`, `tols` and `u`. For example, in our case we are going to pass our simulation model in as an additional argument called `model` (see below).

> **Pre-compilation**: For speed, the simulation models are coded using C/C++ (specifically through the use of the wonderful [`Rcpp`](https://cran.r-project.org/web/packages/Rcpp/index.html) package). This requires that the code is *compiled* into a binary file before running. The `run()` function does this automatically. However, there is an overhead attached with this, and since the ABC-SMC function needs to run the model lots of times, if we used the `run()` function then we would have to compile the model lots of times, which is very costly.
> 
> Instead, we will use a function in `SimBIID` called `compileRcpp()`, which pre-compiles the model into an R function that we can call within our simulation code without having to re-compile the model each time. Hence, to speed things up here we will not use the `run()` function directly, rather we will create the model using `mparseRcpp()`, and then compile it using `compileRcpp()`, before passing the compiled model to the `ABCSMC()` function.

`r ifelse(opts_knit$get("rmarkdown.pandoc.to") == "latex", "\\newpage", "")`

Hence, to compile the model we can run:

```{r, abc-compmod}
model <- compileRcpp(model)
model
```

You can see that the compiled model is an R `function` that takes four arguments, `pars`, `tstart`, `tstop` and `u`. Convenient eh?

Now we generate a function that runs the simulations, extracts the relevant summary statistics, and returns the correct output. Notice that we pass an additional argument `model` that contains our simulation function generated above.

```{r, abc-func}
## set up function to perform simulation and check matching
simBS <- function(pars, data, tols, u, model) {
    
    ## run model
    sims <- model(pars, 0, data[2] + tols[2], u)
    
    ## 'sims' is a vector of outputs of the form:
    ## completed (1/0), t, S, I, B, C
    if(sims[1] == 0) {
        ## if epidemic still going at 
        ## time = finaltime + tol, then reject
        return(NA)
    } else {
        ## extract finaltime and finalsize
        finaltime <- sims[2]
        finalsize <- sims[6]
    }
    
    ## return vector if match, else return NA
    if(all(abs(c(finalsize, finaltime) - data) <= tols)){
        return(c(finalsize, finaltime))
    } else {
        return(NA)
    }
}
```

> **Note**: You can also use you own simulation code in this function, as long as you adhere to these rules, you do not have to use `mparseRcpp()` (you could use the `SimInf` package here for example).

The `simBS` function we have made therefore returns an `NA` if the simulations do not match the summary statistics, or a vector of simulated summary statistics if it does. 

### Initial states {#inistates}

Next we need to specify the initial states of the system (as a named `vector`):

```{r, abc-inistates}
## define initial states
iniStates <- c(S = 762, I = 1, B = 0, C = 0)
```

### Tolerances

We can specify the tolerances in one of two ways:

* either as a `matrix` of tolerances to match to (through the `tols` argument). Note that these must be decreasing over each generation of the ABC-SMC algorithm).
* Alternatively, we can specify the proportion of simulated outcomes at each generation to use to derive adaptive tolerances (through the `ptols` argument). For example, if `ptols = 0.5`, then we choose tolerances at the $t$^th^ generation such that we would have accepted 50% of the simulations at the $(t - 1)$^th^ generation. Note that if we wish to use this latter approach, then we will need to give the algorithm some initial tolerances for generation 1 through the `tols` argument.

> **Note**: the names and order of the tolerances must match the `x` argument.

We will use the second approach, so need to specify a set of tolerances for the first generation only:

```{r, abc-tols}
## set tolerances
tols <- c(finalsize = 50, finaltime = 50)
```

## Running `ABCSMC()`

Now let's run the routine for `ngen = 4` generations of ABC, with `ptols = 0.5` using `npart = 50` particles. Notice that we can pass the additional arguments that `simBS()` needs (`model` in this case) as additional arguments passed to `ABCSMC()`:

```{r, abc-abc}
## run ABC-SMC algorithm
post <- ABCSMC(
    x = sumStat, 
    priors = priors, 
    func = simBS, 
    u = iniStates, 
    tols = tols, 
    ptol = 0.5,
    ngen = 4,
    npart = 50,
    model = model
)
```

 > **Note**: these algorithms are computationally intensive, so if you can spare the processing power, it's often good to set `parallel = T` to use parallel processing. However, depending on the model and your architecture, it's not always faster, so you might need to use your judgement. Windows users commiserate: your architecture doesn't support the `mclapply()` function (from the `parallel` package) so you'll always have to run in serial I'm afraid...

`r ifelse(opts_knit$get("rmarkdown.pandoc.to") == "latex", "\\newpage", "")`

We can plot the approximate posterior distributions over the generations as follows:

```{r, abc-plot, fig.width = 8, fig.height = 4}
## plot approximate posteriors
plot(post)
```

We can plot the simulated summary statistics from the accepted particles at each generation as:

```{r, abc-plot1, fig.width = 8, fig.height = 4}
plot(post, type = "output")
```

Given the speed of the simulations, let's try running for a few more generations. We can pass the current `ABCSMC` object (`post` here) back into the `ABCSMC()` function and run for a few more generations:

```{r, abc-abc1, fig.width = 8, fig.height = 4, eval = F}
## run for a few more generations
post <- ABCSMC(x = post, ptols = 0.5, ngen = 3, parallel = T)
```

`r ifelse(opts_knit$get("rmarkdown.pandoc.to") == "latex", "\\newpage", "")`

```{r, abc-abc11, fig.width = 8, fig.height = 4, echo = F}
## run for a few more generations
post <- ABCSMC(x = post, ptols = 0.5, ngen = 3, parallel = T)
```

```{r, abc-plot2, fig.width = 8, fig.height = 4}
## plot approximate posteriors and simulated
## summary statistics
plot(post)
plot(post, "output")
```

Notice how the approximate posteriors are tightening up as we require to match more precisely. Notice also that the acceptance rate of each generation of ABC-SMC drops as the tolerances decrease. In fact we can see this more clearly if we plot just the first and final generation. Fortunately, we can choose which generations to plot using the `gen` argument to `plot()` e.g.

```{r, abc-plot3, fig.width = 8, fig.height = 4}
## plot approximate posteriors
plot(post, gen = c(1, 7))

## plot accepted outputs
plot(post, "output", gen = c(1, 7))
```

We can also produce joint density plots (if the `GGally` package is installed):

```{r, abc-plot4, fig.width = 6, fig.height = 5, message = F, warning = F}
## joint distributions
plot(post, gen = c(1, 7), joint = T)
```

A call to `summary()` produces weighted estimates for the posterior means and standard deviations for the parameters in the final generation, along with an estimate of the**effective sample size**:

```{r, abc-sum}
summary(post)
```

## Calculate posterior for transformations of parameters

We might also want to calculate posterior distributions for transformations of the parameters. This can be done by specifying a `transfunc` argument to the `summary()` method (this function must return a `data.frame` object, and the arguments to the function must match to names of the parameters used in the transformation). For example, as discussed in the lecture, if we have $\beta$ and $\mu_I$, then we can also calculate the marginal posterior distribution for $R_0$ (here defined as $R_0 = \frac{\beta}{\mu_I}$), as well as, let's say, the length of the infectious period. Hence we must define a function with arguments `beta` and `muI` (which match to parameters in the model), and returns a `data.frame` of transformed parameters:

`r ifelse(opts_knit$get("rmarkdown.pandoc.to") == "latex", "\\newpage", "")`

```{r, abc-R0}
## function to calculate R0 and infectious period
R0fn <- function(beta, muI) {
    data.frame(R0 = beta / muI, infperiod = 1 / muI)
}
summary(post, transfunc = R0fn)
```

We can plot distributions for transformed variables in a similar way:

```{r, abc-plot5, fig.width = 8, fig.height = 4}
## plot distribution for R0
plot(post, transfunc = R0fn, gen = c(1, 7))
```

Note that we can also ask various probabilistic questions of these distributions, since we are using a Bayesian framework. For example, what is the probability that $R_0$ is greater than one? A Monte Carlo estimate of this can be produced simply by counting the proportion of the posterior samples for $R_0$ that lie above one. (**Note**: this would be slightly *biased*, since we actually have unequal weights for these samples. However, we can use our transformation function above to generate an estimate as below.)

```{r, abc-R01}
## function to calculate whether R0 > 1
R0gt <- function(beta, muI) {
    data.frame(R0gt = as.numeric((beta / muI) > 1))
}
summary(post, transfunc = R0gt)
```

## Adding additional matching criteria

Matching to final epidemic size and the date of the final removal only is quite crude. We could also incorporate information regarding the observed $B$ and $C$ curves. One option would be to generate some summary measure of distance between the simulated and observed curves, something like a sum-of-squared differences e.g.
$$
    S_B = \sum_{t = 1}^T \left(B_t - B^\prime_t\right)^2,
$$
where $B_t$ is the **observed** counts at time $t$, and $B^\prime_t$ are the corresponding **simulated** counts at time $t$, and $T$ is the total number of time points. Therefore if the simulated time-series matches the data *exactly*, then $S_B = 0$, with large values of $S_B$ corresponding to poor matches between the simulated and observed counts.

To calculate $S_B$, we have to re-compile the model with the `tspan` option turned on.

```{r, abc-model2}
## specify model
model <- mparseRcpp(
    transitions = transitions, 
    compartments = compartments,
    pars = pars,
    tspan = T
)
## compile into R function
model <- compileRcpp(model)
model
```

You can now see that the new function has an additional `tspan` argument, which corresponds to a vector of times that we wish to return the states of the system. We therefore also have to pass a `tspan` argument to our `simBS` function. In addition, we will also need to pass the observed time-series counts to this function (which we call `B` here).

`r ifelse(opts_knit$get("rmarkdown.pandoc.to") == "latex", "\\newpage", "")`

```{r, abc-func2}
## set up function to perform simulation and check matching
simBS <- function(pars, data, tols, u, model, tspan, B) {
    
    ## run model
    sims <- model(pars, 0, data[2] + tols[2], u, tspan)
    
    ## 'sims' is now a list, with the first element a vector
    ## of the form: completed (1/0), t, S, I, B, C,
    ## and the second element is a matrix containing the 
    ## time-series counts
    simSum <- sims[[1]]
    counts <- sims[[2]]
    if(simSum[1] == 0) {
        ## if epidemic still going at 
        ## time = finaltime + tol, then reject
        return(NA)
    } else {
        ## extract finaltime and finalsize
        finaltime <- simSum[2]
        finalsize <- simSum[6]
        
        ## calculate sum-of-squared distances of
        ## simulations from observations
        SB <- sum((B - counts[, 4])^2)
    }
    
    ## return vector if match, else return NA
    if(all(abs(c(finalsize, finaltime, SB) - data) <= tols)){
        return(c(finalsize, finaltime, SB))
    } else {
        return(NA)
    }
}
```

To create some initial tolerances, let's append an initial value for `SB` to our original `tols` object:

```{r, abc-tols2}
## set up new set of tolerances
tols <- c(tols, SB = 1e6)
```

In practice, we cannot derive a measure of $S_B$ from the **observed** data alone, but we know that if the simulations match the observed data exactly, then $S_B = 0$, so we use this as our target, and append to the `sumStat` vector accordingly:

```{r, abc-tols21}
## set up data to match to
sumStat <- c(sumStat, SB = 0)
```

`r ifelse(opts_knit$get("rmarkdown.pandoc.to") == "latex", "\\newpage", "")`

Now we can run our routine again (notice the final row below contains the additional arguments we now need for `simBS()`):

```{r, abc-abc3}
## run ABC-SMC algorithm
post <- ABCSMC(
    x = sumStat, 
    priors = priors, 
    func = simBS, 
    u = iniStates, 
    npart = 50,
    tols = tols,
    ptols = 0.5,
    ngen = 3, 
    parallel = T,
    model = model, tspan = BS$day, B = BS$B
)
```

```{r, abc-plot6, fig.height = 4, fig.width = 8}
## plot approximate posteriors
plot(post)
```

`r ifelse(opts_knit$get("rmarkdown.pandoc.to") == "latex", "\\newpage", "")`

```{r, abc-plot61, fig.height = 4, fig.width = 8}
## plot simulated summary measures
plot(post, "output")
```

```{task}
Add another summary measure corresponding to the sum-of-squared distances between the observed and simulated $C$ curves. Re-run your ABC-SMC routine. 
```

```{solution}

``{r, abc-solfunc}
## set up function to perform simulation and check matching
simBS <- function(pars, data, tols, u, model, tspan, B, C) {
    
    ## run model
    sims <- model(pars, 0, data[2] + tols[2], u, tspan)
    
    ## 'sims' is now a list, with the first element a vector
    ## of the form: completed (1/0), t, S, I, B, C (here),
    ## and the second element is a matrix containing the 
    ## time-series counts
    simSum <- sims[[1]]
    counts <- sims[[2]]
    if(simSum[1] == 0) {
        ## if epidemic still going at 
        ## time = finaltime + tol, then reject
        return(NA)
    } else {
        ## extract finaltime and finalsize
        finaltime <- simSum[2]
        finalsize <- simSum[6]
        
        ## calculate sum-of-squared distances of
        ## simulations from observations
        SB <- sum((B - counts[, 4])^2)
        SC <- sum((C - counts[, 5])^2)
    }
    
    ## return vector if match, else return NA
    if(all(abs(c(finalsize, finaltime, SB, SC) - data) <= tols)){
        return(c(finalsize, finaltime, SB, SC))
    } else {
        return(NA)
    }
}
``

To create a sequence of tolerances:

``{r, abc-soltol}
## set initial tolerances
tols <- c(tols, SC = 1e6)
``

Finally set up the data to match to and run the model:

``{r, abc-solabc}
## set up data to match to
sumStat <- c(sumStat, SC = 0)

## run ABC-SMC algorithm
post <- ABCSMC(
    x = sumStat, 
    priors = priors, 
    func = simBS, 
    u = iniStates, 
    npart = 50,
    tols = tols,
    ptols = 0.5,
    ngen = 3,
    parallel = T, 
    model = model, tspan = BS$day, B = BS$B, C = BS$C
)
post
summary(post)
``

```

> **Remember: here we have used a low number of particles to speed things up. In practice you would want to use many more to ensure better approximations.**
