# Example: Abakaliki smallpox outbreak

Here we will explore fitting a simple $SIR$ model to the famous Abakaliki smallpox data [see e.g. @oneill_roberts:1999]. This chapter is mainly for you to have a go at!

Firstly, load the `SimBIID` library:

```{r, smallpox-simbiid, warning = F, message = F}
## load library
library(SimBIID)
```

Then load in the data:

```{r, smallpox-data}
## read in the data and convert to vector
smallpox <- as.numeric(unlist(read.csv("smallpox.csv", header = F)))
summary(smallpox)
```
Note that these data, originally from [@thompson_foege:1968], can be found in the `outbreaks` package, albeit in a slightly different form. We use the simpler data in [@oneill_roberts:1999], which correspond to the inter-removal times for `r length(smallpox)` individuals, where the times have been relabelled such that time 0 is the time of the first removal.

## SIR model

Recall, for a continuous-time $SIR$ model we have rates of transition between states given by:
\begin{align*}
    P\left[S_{t + \delta t} = S_t - 1, I_{t + \delta t} = I_t + 1\right] &\approx \beta S I\\
    P\left[I_{t + \delta t} = I_t - 1, R_{t + \delta t} = R_t + 1\right] &\approx \gamma I\\
\end{align*}

## Summary statistics

In the first instance, as before, let's match to:

* final epidemic size (i.e. total number of removals across the time course of the epidemic), and
* time of final removal (in this case when the epidemic process ceased).

Although simple, these two measures serve to give us some information on both the *length* and *magnitude* of the epidemic, and should contain useful information about the parameters. In this case the final removal time is `r max(smallpox)` days and the final epidemic size is `r length(smallpox)` individuals.

## Simulation model

In order to use the `ABCSMC()` function, we need to define a function that runs the simulations and returns an `NA` if the simulation is rejected, or a vector of summary statistics if it is not.

> **Note**: one thing to be careful of for this particular system, is that the data are aligned to time 0 being the time of the first *removal*, not *infection*. Just for illustration, we are going to simplify things in the first instance, and assume that the first infection happens 10 days before the first removal, which corresponds to adding 10 days to each of our observed removal times.

Copy-and-paste the function below. Go through this function and understand what each line is doing.

```{r, smallpox-setup}
## reset times so that initial infection occurs
## 10 days before first removal
smallpox <- smallpox + 10

## define the targeted summary statistics
data <- c(
    finalsize = length(smallpox), 
    finaltime = max(smallpox)
)

## set initial states (1 initial infection in population of 120)
iniStates <- c(S = 119, I = 1, R = 0)
```

```{task}
Set up an SIR model using `mparseRcpp()` to simulate smallpox outbreaks. Check that the model compiles.
```

```{solution}

``{r, smallpox-solsetup}
## set up models
transitions <- c(
    "S -> beta * S * I -> I", 
    "I -> gamma * I -> R"
)
compartments <- c("S", "I", "R")
pars <- c("beta", "gamma")
model <- mparseRcpp(
    transitions = transitions, 
    compartments = compartments,
    pars = pars
)
## compile model for use in ABC
model <- compileRcpp(model)
model
``

```

```{task}
Write a function called `simSIR` that is used to simulate an outbreak and extract the relevant summary statistics from the model runs. the function needs to take at least the arguments: `pars`, `data`, `tols`, `u`, and return an `NA` if no match, or a vector of summary statistics if there is a match.
```

```{solution}

``{r,smallpox-solfunc}
    
## set up function to perform simulation
simSIR <- function(pars, data, tols, u, model) {
    ## run model
    sims <- model(pars, 0, data[2] + tols[2], u)
    
    ## this returns a vector of the form:
    ## completed (1/0), t, S, I, R (here)
    if(sims[1] == 0) {
        ## if simulation rejected
        return(NA)
    } else {
        ## extract finaltime and finalsize
        finaltime <- sims[2]
        finalsize <- sims[5]
    }
    
    ## return vector if match, else return NA
    if(all(abs(c(finalsize, finaltime) - data) <= tols)){
        return(c(finalsize, finaltime))
    } else {
        return(NA)
    }
}
``

```

```{task}
Set up a set of gamma prior distributions for $\beta$ and $\gamma$, with shape parameter 10 in both cases, but rate parameters of 10,000 and 100 respectively.
```

```{solution}

``{r, smallpox-solpriors}
## set priors
priors <- data.frame(
    parnames = c("beta", "gamma"), 
    dist = rep("gamma", 2), 
    stringsAsFactors = F
)
priors$p1 <- c(10, 10)
priors$p2 <- c(10^4, 10^2)
``

```

```{task}
Run 3 generations of the sequential ABC routine of [@tonietal:2009] using starting tolerances of $\epsilon = 50$ and `ptol = 0.2`.
```

```{solution}

``{r, smallpox-solabc}
## set tolerances
tols <- c(
    finalsize = 50,
    finaltime = 50
)

## run ABC-SMC algorithm
post <- ABCSMC(
    x = data, 
    priors = priors, 
    func = simSIR, 
    u = iniStates, 
    tols = tols, 
    ptol = 0.2, 
    ngen = 3, 
    npart = 50,
    parallel = T, 
    model = model
)
``

```

```{task}
Plot the approximate posteriors and simulated outcomes. Produce posterior mean and SD estimates for the parameters.
```

```{solution}

``{r, smallpox-solplot, fig.width = 10, fig.height = 5}
## plot approximate posteriors
plot(post)

## plot accepted outputs
plot(post, "output")

## summarise posteriors
summary(post)
``

```
