# More challenging example: Ebola in the Democratic Republic of Congo

The rates of transition between states in this model is given by:

\begin{align*}
    P\left[\mbox{infection event in}~[t, t + \delta t)\right] = \beta S I / N + o(\delta t)\\
    P\left[\mbox{infectivity event in}~[t, t + \delta t)\right] = \delta E + o(\delta t)\\
    P\left[\mbox{removal event in}~[t, t + \delta t)\right] = \gamma I + o(\delta t)
\end{align*}

where $\beta$ is *time-dependent*, such that:

$$
    \beta = \left\{
    \begin{align}
        &\beta & \mathrm{if}~t < t_{int},\\
        &\beta e^{-q(t - t_{int})} & \mathrm{if}~t \geq t_{int},
    \end{align}\right.
$$

where $t_{int}$ is the time at which intervention strategies were introduced. The challenge here is that in a stochastic system we require that the event rates are piecewise constant, and thus the continuous exponential decay is hard to capture. Instead we can approximate this by assuming that the rate only decays at the beginning of each day. Hence the transmission term becomes:

$$
    \beta = \left\{
    \begin{align}
        &\beta & \mathrm{if}~t < t_{int},\\
        &\beta e^{-q(\lfloor t \rfloor - t_{int})} & \mathrm{if}~t \geq t_{int},
    \end{align}\right.
$$
where $\lfloor t \rfloor$ corresponds to rounding $t$ down to the nearest integer. Here $t_{int} = 126$, meaning that control interventions were introduced on day 126 of the outbreak. Hence the transmission rate between day 126 and 127 is:
$$
    \beta e^{-q(126 - 126)} = \beta,
$$
the transmission rate between day 127 and 128 is:
$$
    \beta e^{-q(127 - 126)} = \beta e^{-q},
$$
and so on. This behaviour can be captured in C by using the `floor()` function (which rounds down to the nearest integer), and an **if-else** statement, which in C is coded as `(CONDITION ? X:Y)`, where `CONDITION` is some conditional statement, which if **true** returns `X` and if **false** returns `Y`. Hence we can define our transmission rate to be:

`beta * exp(-q * (t > 126.0 ? (floor(t) - 126.0):0.0))`

This says that at time `t`, if `t` is **less than** $t_{int}$ then the rate is $\beta e^{-q \times 0} = \beta$, whereas if `t` is **greater than** $t_{int}$ then the transmission rate is $\beta e^{-q (\lfloor t \rfloor - 126)}$ as required. Now we can set up our model:

```{r}
## set up model
transitions <- c(
    "S -> beta * exp(-q * (t > 126.0 ? (floor(t) - 126.0):0.0)) * S * I / (S + E + I + R) -> E", 
    "E -> delta * E -> I", 
    "I -> gamma * I -> R"
)
compartments <- c("S", "E", "I", "R")
pars <- c("beta", "delta", "gamma", "q")
model <- mparseRcpp(
    transitions = transitions, 
    compartments = compartments,
    pars = pars
)
model <- compileRcpp(model)
model
```

Now we can set priors and initial states:

```{r}
## set priors
priors <- data.frame(
    parnames = c("beta", "delta", "gamma", "q"), 
    dist = rep("gamma", 4), 
    stringsAsFactors = F
)
priors$p1 <- rep(2, 4)
priors$p2 <- c(10, 10, 1 / 0.07, 10)

## set initial states
iniStates <- data.frame(S = 5364499, E = 0, I = 1, R = 0)
```

Now we set up the data to match to (final epidemic size and date of final removal):

```{r}
## define the targeted summary statistics
data <- data.frame(finalsize = 316, finaltime = 191)
```

Finally we set the number of particles and a sequence of tolerances:

```{r}
## set number of particles required
npart <- 200

## set tolerances
tols <- matrix(rep(round(seq(300, 40, length.out = 10)), each = 2), ncol = 2, byrow = T)
tols <- as.data.frame(tols)
colnames(tols) <- c("finalsize", "finaltime")
```

We then set up a function to run the model, extract the final epidemic size and date of the final removal, and return the relevant measures:

```{r}
## function to match simulations
simEbola <- function(pars, data, tols, u, model) {
    ## run model
    sims <- model(pars, 0, data[2] + tols[2], u)
    
    ## this returns a vector of the form:
    ## completed (1/0), t, S, E, I, R (here)
    if(sims[1] == 0) {
        ## if simulation rejected
        return(NA)
    } else {
        ## extract finaltime and finalsize
        finaltime <- sims[2]
        finalsize <- sims[6]
    }
    
    ## return vector if match, else return NA
    if(all(abs(c(finalsize, finaltime) - data) <= tols)){
        return(c(finalsize, finaltime))
    } else {
        return(NA)
    }
}
```

Now we run the ABC-SMC algorithm:

```{r}
## run ABC-SMC algorithm
post <- ABCSMC(data, priors, simEbola, iniStates, npart, tols, parallel = T, model = model)
```

```{r}
## function to calculate R0 and length of
## epidemiological periods
R0fn <- function(beta, delta, gamma) {
    data.frame(R0 = beta / gamma, latperiod = 1 / delta, infperiod = 1 / gamma)
}

## summarise approximate posterior
summary(post, transfunc = R0fn)
```

```{r, fig.width = 8, fig.height = 8}
## plot approximate posteriors
plot(post, transfunc = R0fn, gen = c(1, 5, 10))
```

```{r}
## plot accepted outputs
plot(post, "output", gen = c(1, 5, 10))
```

This ran relatively quickly, so now run for another few generations...

```{r}
tols <- matrix(rep(seq(30, 10, by = -10), each = 2), ncol = 2, byrow = T)
tols <- as.data.frame(tols)
colnames(tols) <- c("finalsize", "finaltime")

## run ABC-SMC algorithm
post <- ABCSMC(post, tols, parallel = T)
```

Now summarise the outputs:

```{r, fig.width = 10, fig.height = 5}
## summarise approximate posterior
summary(post, transfunc = R0fn)
```

```{r, fig.width = 8, fig.height = 8}
## plot approximate posteriors
plot(post, transfunc = R0fn, gen = c(1, 5, 13))
```

```{r}
## plot accepted outputs
plot(post, "output", gen = c(1, 5, 13))
```
