# Set up simple simulation model

This chapter is to illustrate how to use `mparseRcpp` to simulate from the flu model that we wish to fit to the boarding school data.

```{r, warning = F, message = F}
## load library
library(SimBIID)
```

To set up the model above we need a character vector of transition rates:

```{r}
## test compilation model
transitions <- c(
    "S -> beta * S * I / (S + I + B + C) -> I", 
    "I -> muI * I -> B", 
    "B -> muB * B -> C"
)
```

We also need a vector of compartment names and parameter names:

```{r}
compartments <- c("S", "I", "B", "C")
pars <- c("beta", "muI", "muB")
```

We then build a model using the `mparseRcpp()`:

```{r}
model <- mparseRcpp(
    transitions = transitions, 
    compartments = compartments,
    pars = pars
)
```

The `SimBIID` package provides a `run()` method to run individual or multiple simulations from the model. By default the `mparseRcpp` creates a function that takes four arguments:

* `pars`: a vector of parameter values;
* `tstart`: the time to begin the simulation;
* `tstop`: the time to end the simulation;
* `u`: states of the system at `tstart`.

We can then pass this object to the `run()` method, along with the corresponding arguments e.g.

```{r}
## define parameters
simPars <- data.frame(beta = 3.7, muI = 2.2, muB = 0.78)

## define initial states
iniStates <- data.frame(S = 762, I = 1, B = 0, C = 0)

## run model
sims <- run(
    model,
    pars = simPars,
    tstart = 0,
    tstop = 14,
    u = iniStates
)
sims
```

Currently this outputs a `data.frame` with columns of the form: `completed / t / u*`, where `completed` is a binary variable containing the value `1` if the epidemic finished before time point `tstop`, and `0` otherwise. Then the value `t` is the time point when the epidemic finished (or `tstop` if the epidemic was still ongoing when the function exited), and `u*` corresponds to a copy of the states of the system at the final time point.

Here we can see that the simulation finished at time point `t = 13.95`, with the final state of the system given by `c(S = 192, I = 0, B = 0, C = 571)`.

### `tspan`

Alternatively, we may want to return the states of the system at a series of pre-determined points. We can do this by adding a `tspan` argument, which is a vector of time points at which we wish to return the states. We initialise the model by telling `mparseRcpp` that we wish to include a `tspan` argument e.g.

```{r}
## generate model
model <- mparseRcpp(
    transitions = transitions, 
    compartments = compartments,
    pars = pars,
    tspan = T
)
```

When we `run()` the model, we can enter a suitable `tspan` argument as a vector of time points at which we wish to return the states e.g.

```{r, include = F}
set.seed(10)
```

```{r}
sims <- run(
    model,
    pars = simPars,
    tstart = 0,
    tstop = 14,
    u = iniStates,
    tspan = 1:14
)
sims
```

Here we can see that the epidemic is still going at time point `t = 14`, but that there are no more infectives left, so the epidemic will complete shortly (once the remaining individual in the $B$ class moves to the $R$ class).

### Plotting simulations

The `run()` method outputs a `SimBIID_runs` object, which can be plotted using the `plot()` function:

```{r}
## plot simulation
plot(sims)
```

### Running multiple simulations

We can run multiple simulations by passing a `nrep` argument to `run()`. For example, to run 100 replicates and plot them:

```{r}
sims <- run(
    model,
    pars = simPars,
    tstart = 0,
    tstop = 14,
    u = iniStates,
    tspan = 1:14,
    nrep = 100
)
sims
plot(sims)
```

We can add individual trajectories by inputing their replicate number, e.g.

```{r}
plot(sims, rep = c(1, 2))
```

There is an argument in `run()` to enable parallel processing if required (if the `parallel` package is installed). See help file for `run()` e.g. `?run` for more details.
